---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document: 
    keep_md: true
---


## Loading and preprocessing the data
### load libraries
```{r echo=TRUE}
library(tidyverse)
library(dplyr)
library(gridExtra)
```

Read CSV datafiles & convert date strings to dates 
```{r echo=TRUE}
PAdata <- read.csv("activity.csv", stringsAsFactors = FALSE)
PAdata$date <- as.Date(PAdata$date, "%Y-%m-%d")
summary(PAdata)
```

Group by date & summarize
```{r echo=TRUE}
PAdata_bg <- group_by(PAdata, date)
PAdata_summary <- summarize(PAdata_bg, totalSteps=sum(steps, na.rm=TRUE))
```

## What is mean total number of steps taken per day?
### plot histogram
```{r echo=TRUE}
hist(PAdata_summary$totalSteps, 
     xlab="Number of Steps", 
     main="Histogram for Number of Steps per Day",
     breaks=15, col="lightblue")
```

### Summarize data grouped by date to get mean and median
```{r echo=TRUE}
s1 <- summarize(PAdata_bg, mean(steps, na.rm=TRUE), median(steps, na.rm=TRUE))
names(s1) <- c("date", "means", "medians")
print(s1)
```

## What is the average daily activity pattern?
### Plot time series of averages
```{r Pecho=TRUE}
p2 <- ggplot(s1, aes(x=date, y=means, na.rm=TRUE )) + 
  geom_bar(stat="identity", color="dark blue", fill="steelblue4") +
  ggtitle("Average Number of Steps - Day")
print(p2)
```


## Imputing missing values
### impute for NA values
``` {r echo=TRUE}
totalNA <- sum(complete.cases(PAdata))
print(totalNA)
```

Substitute NAs with value - means from same interval
``` {r echo=TRUE}
averageSteps <- aggregate(steps ~ interval, PAdata, FUN=mean)
newSet <- numeric()
for (i in 1:nrow(PAdata)) {
  if(is.na(PAdata[i,]$steps)) {
    steps <- subset(averageSteps, interval == PAdata[i,]$interval)$steps
  }
  else
  {
    steps <- PAdata[i,]$steps
  }
  newSet <- c(newSet, steps)
}
```

Create new dataset with NAs substituted
``` {r echo=TRUE}
PAdata_new <- PAdata
PAdata_new$steps <- newSet
```


## Are there differences in activity patterns between weekdays and weekends?

### Panel Graph for weekday and weekend average steps by Interval
```{r echo=TRUE}
PAdata_int <- PAdata %>% mutate(DayType=ifelse(weekdays(date)=="Saturday" |
                                                 weekdays(date)=="Sunday",'Weekend','Weekday'))
PAdata_summ <- PAdata_int %>% 
  group_by(interval, DayType) %>% 
  summarize(AverageSteps=mean(steps, na.rm=TRUE))
par(mai=c(1,0.8,0.5,0.5), mfrow=c(2,1))

with(filter(PAdata_summ, DayType=='Weekend'), 
     plot(x=interval, y=AverageSteps, type="l", main="Weekend", ylab="", cex.main=0.9, xlab=""))
with(filter(PAdata_summ, DayType=='Weekday'), 
     plot(x=interval, y=AverageSteps, type="l", main="Weekday", ylab="", cex.main=0.9, xlab="Interval"))

```

